Gravitational wave sources in our Galactic backyard: Predictions for BH and NS binaries in LISA

### 2021-07-23 paper/tex/sections/conclusion.tex
ensure all stats are up to date

Our main conclusions can be summarised as:
\begin{itemize}
    \item \textbf{Detection rate:} We find that on average, for our fiducial model, a 4(10)-year LISA mission will detect \BHBHFourYear{}(\BHBHTenYear{}) BHBHs, \BHNSFourYear{}(\BHNSTenYear{}) BHNSs and \NSNSFourYear{}(\NSNSTenYear{}) NSNSs.
    \item \textbf{Black hole mass distribution:} We find that the black hole mass distribution for both BHBHs and BHNSs that are detectable with LISA are not skewed towards heavier masses (as in ground-based detectors), with $\BHBHmBHBelowEleven{}$ and $\BHNSmBHBelowEleven{}$ having masses less than $11 \unit{M_\odot}$ respectively. Moreover, we find that for LISA detectable DCOs, $\BHBHLowerMassGap{}$ of BHBHs and $\BHNSLowerMassGap{}$ of BHNSs have masses in the theoretical lower mass gap between NSs and BHs. This implies that LISA will be able provide conclusive evidence about the existence of a lower-mass gap. 
    \item \textbf{Eccentricity distribution:} We find that BH and NS binaries still have significant eccentricities when observable by the LISA mission, unlike the more numerous WDWD LISA population. We find that \BHBHNotCirc{}(\BHBHHighlyEccentric{}) of BHBHs, \BHNSNotCirc{}(\BHNSHighlyEccentric{}) of BHNSs and \NSNSNotCirc{}(\NSNSHighlyEccentric{})\% of NSNSs have eccentricity of $e > 0.01(0.3)$ at the start of the LISA mission.
    \item \textbf{Physics variations:} For BHBHs, we find that the detection rate is largely unaffected by changes in underlying physics assumptions, except for the optimistic CE scenario and increased Wolf-Rayet winds. Conversely, the BHNS rate varies widely with physics variation, ranging across three orders of magnitude. The NSNS shows some variation but only shows large changes under the assumption that case BB mass transfer is unstable and that core-collapse supernovae produce weaker kicks.
    \item \textbf{Distinguishable sources:} For a 4(10)-year LISA mission, we estimate that of the detectable population, at least \BHBHNotWDWDFour{}(\BHBHNotWDWDTen{}) BHBHs, \BHNSNotWDWDFour{}(\BHNSNotWDWDTen{}) BHNSs and \NSNSNotWDWDFour{}(\NSNSNotWDWDTen{}) NSNSs will produce signals that are distinguishable from a signal produced by a WDWD. Additionally, we predict that we will be able to determine whether \BHBHDistinguishedFour{}(\BHBHDistinguishedTen{}) BHBHs, \BHNSDistinguishedFour{}(\BHNSDistinguishedTen{}) BHNSs and \NSNSDistinguishedFour{}(\NSNSDistinguishedTen{}) NSNSs are signals from a binary that contains a black hole.
    \item \textbf{Joint SKA-LISA detections:} We expect that \textit{if} the binary contains a pulsar that is beaming towards Earth, SKA-1 will be able to detect at least 10 LISA DCOs that contain a NS, whilst the increased number of pulsars that could crowd the sky in SKA-2 means that this total decreases to 6. This total could be much higher if we consider that SKA and LISA could match their orbital frequency estimates to select the correct pulsar from a crowded field.
    \item \textbf{Importance of choice of MW model:} Many studies use Milky Way models that use fixed metallicity populations which are assigned irrespective of birth time or position, do not account for the inside-out growth of the thin disc and use constant star formation rates. The use of these simpler MW models may lead to an overestimation of the LISA NSNS detection rate. It may also introduce unphysical artifacts into DCO parameter distributions, particularly the mass distributions, which lead to inaccurate predictions.
\end{itemize}

>>>>

Our main conclusions can be summarised as:
\begin{itemize}
    \item \textbf{Detection rate:} We find that on average, for our fiducial model, a 4(10)-year LISA mission will detect \BHBHFourYear{}(\BHBHTenYear{}) BHBHs, \BHNSFourYear{}(\BHNSTenYear{}) BHNSs and \NSNSFourYear{}(\NSNSTenYear{}) NSNSs.
    \item \textbf{Black hole mass distribution:} We find that the black hole mass distribution for both BHBHs and BHNSs that are detectable with LISA are not skewed towards heavier masses (as in ground-based detectors), with 90\% having BH masses less than $11 \unit{M_\odot}$. Moreover, we find that for LISA detectable DCOs, $\BHBHLowerMassGap{}$ of BHBHs and $\BHNSLowerMassGap{}$ of BHNSs have masses in the theoretical lower mass gap between NSs and BHs. This implies that LISA will be able provide conclusive evidence about the existence of a lower-mass gap. 
    \item \textbf{Eccentricity distribution:} We find that BH and NS binaries still have significant eccentricities when observable by the LISA mission, unlike the more numerous WDWD LISA population. We find that \BHBHNotCirc{}(\BHBHHighlyEccentric{}) of BHBHs, \BHNSNotCirc{}(\BHNSHighlyEccentric{}) of BHNSs and \NSNSNotCirc{}(\NSNSHighlyEccentric{})\% of NSNSs have eccentricity of $e > 0.01(0.3)$ at the start of the LISA mission.
    \item \textbf{Physics variations:} For BHBHs, we find that the detection rate is largely unaffected by changes in underlying physics assumptions, except for the optimistic CE scenario and increased Wolf-Rayet winds. Conversely, the BHNS rate varies widely with physics variation, ranging across three orders of magnitude. The NSNS shows some variation but only shows large changes under the assumption that case BB mass transfer is unstable and that core-collapse supernovae produce weaker kicks.
    \item \textbf{Distinguishable sources:} For a 4(10)-year LISA mission, we estimate that of the detectable population, at least \BHBHNotWDWDFour{}(\BHBHNotWDWDTen{}) BHBHs, \BHNSNotWDWDFour{}(\BHNSNotWDWDTen{}) BHNSs and \NSNSNotWDWDFour{}(\NSNSNotWDWDTen{}) NSNSs will produce signals that are distinguishable from a signal produced by a WDWD. Additionally, we predict that we will be able to determine whether \BHBHEitherBHOrNSFour{}(\BHBHEitherBHOrNSTen{}) BHBHs, \BHNSEitherBHOrNSFour{}(\BHNSEitherBHOrNSTen{}) BHNSs and \NSNSEitherBHOrNSFour{}(\NSNSEitherBHOrNSTen{}) NSNSs are signals from a binary that contains a black hole.
    \item \textbf{Joint SKA-LISA detections:} We expect that \textit{if} the binary contains a pulsar that is beaming towards Earth, SKA-1 will be able to detect at least 10 LISA DCOs that contain a NS, whilst the increased number of pulsars that could crowd the sky in SKA-2 means that this total decreases to 6. This total could be much higher if we consider that SKA and LISA could match their orbital frequency estimates to select the correct pulsar from a crowded field.
    \item \textbf{Importance of choice of MW model:} Many studies use Milky Way models that use fixed metallicity populations which are assigned irrespective of birth time or position, do not account for the inside-out growth of the thin disc and use constant star formation rates. The use of these simpler MW models may lead to an overestimation of the LISA NSNS detection rate. It may also introduce unphysical artifacts into DCO parameter distributions, particularly the mass distributions, which lead to inaccurate predictions.
\end{itemize}

### 2021-07-23 paper/tex/sections/discussion.tex
ensure all stats are up to date

\subsubsection{Distinguishing from WDWD population}\label{sec:WDWD_distinguish}
The simplest way to check whether a source is a WDWD is to check its chirp mass. The mass of a non-rotating white dwarf cannot be larger than the Chandrasekhar limit of $1.44 \unit{M_\odot}$ \citep{Chandrasekhar+1931}, so we can take the maximum chirp mass of a WDWD to be $\sim 1.25 \unit{M_{\odot}}$. Therefore, any DCO with a chirp mass that satisfies $\mathcal{M}_c > 1.25 \unit{M_{\odot}} + 2 \Delta \mathcal{M}_c$ must not be a WDWD. We find that for the detectable population of a 4(10)-year LISA mission, 17(23)\% of BHBHs, 20(24)\% of BHNSs and 2.4(2.5)\% of NSNSs satisfy this condition. This method is not particularly effective for NSNSs since their average chirp mass, $1.17 \unit{M_\odot}$, is below the Chandrasekhar limit.

Another discriminator between WDWDs and other DCOs is the eccentricity. WDWDs formed in the disc are thought to be formed through isolated binary formation and have little to no eccentricity \citep[e.g.][]{Nelemans+2001}. Therefore, if any system is detected with anything other than one detectable harmonic, this implies that the system is unlikely to be a WDWD. We find that for a 4(10)-year LISA mission, 55(61)\% of BHBHs, 27(28)\% of BHNSs and 65(69)\% of NSNSs are detected with multiple harmonics. Both the absolute percentage and the relative improvement with an extended LISA mission is lower for the BHNSs with respect to other DCOs as we find that these BHNSs are less eccentric on average (see Fig.~\ref{fig:fiducial_pdf_distributions} and discussion in Sec. \ref{sec:fiducial_distributions}).

However, we should also consider that eccentric WDWDs could be formed through dynamical formation in Milky Way globular clusters \citep[e.g.][]{Willems+2007, Kremer+2018}. This means that we cannot assume that eccentric binaries are not WDWDs unless they are detected in the Galactic plane. We can use the sky localisation, scale height of the disc and distance to the source to estimate what fraction of eccentric sources can be localised to the Galactic plane. This condition can be written as $\sigma_\theta < \arcsin(z_{\rm plane} / D_L)$ or $D_L < z_{\rm plane}$, where we set the height of the Galactic plane, $z_{\rm plane}=0.95 \unit{kpc}$, to the scale height of the high-$\alpha$ disc. We apply this condition to find that the fraction of sources that are eccentric \textit{and} localised within the disc for a 4(10)-year LISA mission are 40(32)\% for BHBHs, 24(19)\% for BHNSs and 59(51)\% for NSNSs\footnote{Note that although the fractions are smaller for the 10 year mission, the \textit{absolute} number of detections is still greater. The fraction decreases because a 10 year mission detects more `marginal' sources that are just on the cusp of the detection threshold and these sources have the worst sky localisation and thus cannot be confirmed to lie within the Galactic plane.}.

@@

The problem of discriminating between the BHBH, BHNS and NSNS populations can be more difficult than distinguishing them from WDWDs. For NSNSs, we can follow a similar method to the WDWDs (see Sec.~\ref{sec:WDWD_distinguish}) by applying our knowledge of the maximum mass of a neutron star. Following our fiducial assumption, we can take the maximum mass of a neutron star as $2.5 \unit{M_{\odot}}$ and thus the maximum chirp mass that a system can attain without one of the components being a black hole is $\mathcal{M}_{c} = 2.2 \unit{M_\odot}$. For a 4(10)-year LISA mission, the fraction of systems that are above or below this limit by more than $2 \Delta \mathcal{M}_c$ is 15(20)\% for BHBHs, 12(14)\% for BHNSs and 34(40)\% of NSNSs, which in terms of absolute detections is \BHBHDistinguishedFour{}(\BHBHDistinguishedTen{}) for BHBHs, \BHNSDistinguishedFour{}(\BHNSDistinguishedTen{}) for BHNSs and \NSNSDistinguishedFour{}(\NSNSDistinguishedTen{}) for NSNSs.

>>>>

\subsubsection{Distinguishing from WDWD population}\label{sec:WDWD_distinguish}
The simplest way to check whether a source is a WDWD is to check its chirp mass. The mass of a non-rotating white dwarf cannot be larger than the Chandrasekhar limit of $1.4 \unit{M_\odot}$ \citep{Chandrasekhar+1931, Hamada+1961}, so we can take the maximum chirp mass of a WDWD to be $\sim 1.2 \unit{M_{\odot}}$. Therefore, any DCO with a chirp mass that satisfies $\mathcal{M}_c > 1.2 \unit{M_{\odot}} + 2 \Delta \mathcal{M}_c$ must not be a WDWD. We find that for the detectable population of a 4(10)-year LISA mission, \BHBHAboveMaxWDWDFourPerc{}(\BHBHAboveMaxWDWDTenPerc{})\% of BHBHs, \BHNSAboveMaxWDWDFourPerc{}(\BHNSAboveMaxWDWDTenPerc{})\% of BHNSs and \NSNSAboveMaxWDWDFourPerc{}(\NSNSAboveMaxWDWDTenPerc{})\% of NSNSs satisfy this condition. This method is not particularly effective for NSNSs since their average chirp mass, $1.17 \unit{M_\odot}$, is below the Chandrasekhar limit.

Another discriminator between WDWDs and other DCOs is the eccentricity. WDWDs formed in the disc are thought to be formed through isolated binary formation and have little to no eccentricity \citep[e.g.][]{Nelemans+2001}. Therefore, if any system is detected with anything other than one detectable harmonic, this implies that the system is unlikely to be a WDWD. We find that for a 4(10)-year LISA mission, \BHBHMultipleHarmonicsFourPerc{}(\BHBHMultipleHarmonicsTenPerc{})\% of BHBHs, \BHNSMultipleHarmonicsFourPerc{}(\BHNSMultipleHarmonicsTenPerc{})\% of BHNSs and \NSNSMultipleHarmonicsFourPerc(\NSNSMultipleHarmonicsTenPerc{})\% of NSNSs are detected with multiple harmonics. Both the absolute percentage and the relative improvement with an extended LISA mission is lower for the BHNSs with respect to other DCOs as we find that these BHNSs are less eccentric on average (see Fig.~\ref{fig:fiducial_pdf_distributions} and discussion in Sec. \ref{sec:fiducial_distributions}).

However, we should also consider that eccentric WDWDs could be formed through dynamical formation in Milky Way globular clusters \citep[e.g.][]{Willems+2007, Kremer+2018}. This means that we cannot assume that eccentric binaries are not WDWDs unless they are detected in the Galactic plane. We can use the sky localisation, scale height of the disc and distance to the source to estimate what fraction of eccentric sources can be localised to the Galactic plane. This condition can be written as $\sigma_\theta < \arcsin(z_{\rm plane} / D_L)$ or $D_L < z_{\rm plane}$, where we set the height of the Galactic plane, $z_{\rm plane}=0.95 \unit{kpc}$, to the scale height of the high-$\alpha$ disc. We apply this condition to find that the fraction of sources that are eccentric \textit{and} localised within the disc for a 4(10)-year LISA mission are \BHBHEccInDiscFourPerc{}(\BHBHEccInDiscTenPerc{})\% for BHBHs, \BHNSEccInDiscFourPerc{}(\BHNSEccInDiscTenPerc{})\% for BHNSs and \NSNSEccInDiscFourPerc{}(\NSNSEccInDiscTenPerc{})\% for NSNSs. Note that although the fractions are smaller for the 10-year mission, the \textit{absolute} number of detections is still greater. The fraction decreases because a 10-year mission detects more `marginal' sources that are just on the cusp of the detection threshold and these sources have the worst sky localisation and thus cannot be confirmed to lie within the Galactic plane.

@@

The problem of discriminating between the BHBH, BHNS and NSNS populations can be more difficult than distinguishing them from WDWDs. For NSNSs, we can follow a similar method to the WDWDs (see Sec.~\ref{sec:WDWD_distinguish}) by applying our knowledge of the maximum mass of a neutron star. Following our fiducial assumption, we can take the maximum mass of a neutron star as $2.5 \unit{M_{\odot}}$ and thus the maximum chirp mass that a system can attain without one of the components being a black hole is $\mathcal{M}_{c} = 2.2 \unit{M_\odot}$. For a 4(10)-year LISA mission, the fraction of systems that are above or below this limit (and thus \textit{must} contain or not contain a BH component respectively) by more than $2 \Delta \mathcal{M}_c$ is \BHBHEitherBHOrNSFourPerc{}(\BHBHEitherBHOrNSTenPerc{})\% for BHBHs, \BHNSEitherBHOrNSFourPerc{}(\BHNSEitherBHOrNSTenPerc{})\% for BHNSs and \NSNSEitherBHOrNSFourPerc{}(\NSNSEitherBHOrNSTenPerc{})\% of NSNSs, which in terms of absolute detections is \BHBHEitherBHOrNSFour{}(\BHBHEitherBHOrNSTen{}) for BHBHs, \BHNSEitherBHOrNSFour{}(\BHNSEitherBHOrNSTen{}) for BHNSs and \NSNSEitherBHOrNSFour{}(\NSNSEitherBHOrNSTen{}) for NSNSs.

### 2021-07-23 paper/tex/sections/results.tex
ensure all stats are up to date

\paragraph{Black Hole Mass}
For both the BHBHs and BHNSs, the black hole mass distribution extends to relatively low masses, with $\BHBHmBHBelowEleven{}$ and $\BHNSmBHBelowEleven{}$ respectively below $11 \unit{M_{\odot}}$. Unlike ground-based detectors, LISA is not biased to higher masses and so the mass distribution more closely follows the IMF. In addition, at the high metallicities in the Milky Way, stellar winds are much stronger and strip away much of the stellar mass before BH formation, resulting the less massive black holes. The mass distribution extends down to $2.5 \unit{M_{\odot}}$, our fiducial maximum neutron star mass. Since the \citet{Fryer+2012} \textit{delayed} remnant mass prescription (applied in this work) does not produce a mass gap between neutron stars and black holes. Indeed we expect $\BHBHLowerMassGap{}$ and $\BHNSLowerMassGap{}$ of detected BHBH and BHNS systems to contain a black hole in the lower mass gap. During the preparation of this paper \citet{Shao+2021} also noted this, claiming that $30\%$ and $70\%$ of merging BHBHs and BHNSs respectively have BHs in the lower mass gap (when using the delayed remnant mass prescription). Therefore, LISA could help to confirm or rule out the existence of the lower mass gap.

>>>>

\paragraph{Black Hole Mass}
For both the BHBHs and BHNSs, the black hole mass distribution extends to relatively low masses, with about $90\%$ of BHs below $11 \unit{M_{\odot}}$. Unlike ground-based detectors, LISA is not biased to higher masses and so the mass distribution more closely follows the IMF. In addition, at the high metallicities in the Milky Way, stellar winds are much stronger and strip away much of the stellar mass before BH formation, resulting the less massive black holes. The mass distribution extends down to $2.5 \unit{M_{\odot}}$, our fiducial maximum neutron star mass. Since the \citet{Fryer+2012} \textit{delayed} remnant mass prescription (applied in this work) does not produce a mass gap between neutron stars and black holes. Indeed we expect $\BHBHLowerMassGap{}$ and $\BHNSLowerMassGap{}$ of detected BHBH and BHNS systems to contain a black hole in the lower mass gap. During the preparation of this paper \citet{Shao+2021} also noted this, claiming that $30\%$ and $70\%$ of merging BHBHs and BHNSs respectively have BHs in the lower mass gap (when using the delayed remnant mass prescription). Therefore, LISA could help to confirm or rule out the existence of the lower mass gap.
